Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-06-11 paper.qmd
Consistent with the word hyper-parameter

---
title: "Looking at Non-Linear Dimension Reductions as Models in the Data Space"
format: 
    jasa-pdf:
        keep-tex: true
    jasa-html: default
author:
  - name: Jayani P. Gamage
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-6265-6481
    email: jayani.piyadigamage@monash.edu
    url: https://jayanilakshika.netlify.app/
  - name: Dianne Cook
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-3813-7155
    email: dicook@monash.edu 
    url: http://www.dicook.org/
  - name: Paul Harrison
    affiliations:
      - name: Monash University
        department: MGBP, BDInstitute
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-3980-268X
    email: 	paul.harrison@monash.edu
    url: 
  - name: Michael Lydeamore
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0001-6515-827X
    email: michael.lydeamore@monash.edu
    url: https://www.michaellydeamore.com/
  - name: Thiyanga S. Talagala
    affiliations:
      - name: University of Sri Jayewardenepura
        department: Statistics
        address: Gangodawila
        city: Nugegoda 
        country: Sri Lanka
        postal-code: 10100
    orcid: 0000-0002-0656-9789
    email: ttalagala@sjp.ac.lk 
    url: https://thiyanga.netlify.app/
tbl-cap-location: bottom
abstract: |
  Non-linear dimension reduction (NLDR) techniques such as tSNE, and UMAP provide a low-dimensional representation of high-dimensional data (\pD{}) by applying a non-linear transformation. NLDR often exaggerates random patterns, sometimes due to the samples observed. But NLDR views have an important role in data analysis because, if done well, they provide a concise visual (and conceptual) summary of \pD{} distributions. The NLDR methods and (hyper)parameter choices can create wildly different representations, making it difficult to decide which is best, or whether any or all are accurate or misleading. To help assess the NLDR and decide on which, if any, is the most reasonable representation of the structure(s) present in the \pD{} data, we have developed an algorithm to show the \gD{} NLDR model in the \pD{} space, viewed with a tour, a movie of linear projections. From this, one can see if the model fits everywhere, or better in some subspaces, or completely mismatches the data. Also, we can see how different methods may have similar summaries or quirks.

@@

We define the NLDR as a function $g\text{:}~ \mathbb{R}^{n\times p} \rightarrow \mathbb{R}^{n\times k}$, with (hyper-)parameters $\mathbfit{\theta}$. The parameters, $\mathbfit{\theta}$, depend on the choice of $g$, and can be considered part of model fitting in the traditional sense. Common choices for $g$ include functions used in tSNE, UMAP, PHATE, TriMAP, PaCMAP, or MDS, although in theory any function that does this mapping is suitable.

@@

```{r}
#| label: best-fit-tsne
#| fig-cap: "Lifting the fitted model into \\pD{}. Starting from a default tSNE layout (a) of non-linear clusters data ($n =  2000$ and $p = 4$), from which the \\gD{} wireframe is constructed (a) and then lifted to $4\\text{-}D$ (b, c). The fit is reasonbly tight with the data, although it does not fully spread the full width. This highlights a key characteristic of tSNE, which compresses the data when applying tSNE. Furthermore, the sparse space observed in the middle of the tSNE layout is a result of (hyper-)parameter choice. Video of the langevitour animation is available at <>."
#| fig-width: 8
#| fig-height: 2

@@

@fig-five-gau-projs a1, b1, c1 show the \gD{} layouts for (a) tSNE, (b) UMAP, and (c) PaCMAP, respectively. The default hyper-parameters for each method are used. In each layout we can see an accurate representation where all five clusters are visible, although with varying degrees of separation.

@@

@fig-one-dens_clust-error (a2, b2) shows \gD{} layout for tSNE created using the default hyper-parameters. One would expect to see a rectangular shape if the curved sheet is flattened, but the layout is triangular. The other two displays show the residuals as a dot density plot (a1, b1), and a \gD{} projection of the data and the model from \fD{} (a3, b3). Using linked brushing between the plots, we can highlight points in the tSNE layout, and examine where they fall in the original \fD{}. The darker (maroon) points indicate points that have been highlighted by linking. In row a, the points at the top of the triangle are highlighted, and we can see these correspond to higher residuals, and also to all points at the low density end of the curved sheet. In row b, points at the lower left side of the triangle are highlighted which corresponds to smaller residuals and one corner of the sheet at the high density end of the curved sheet.

@@

```{r}
#| echo: false
#| fig-cap: "Compare the published \\gD{} layout (@fig-pbmc-mse a) and the \\gD{} layout selected (@fig-pbmc-mse e) by MSE plot (@fig-pbmc-mse) from the UMAP and tSNE with default hyper-parameters. The PBMC3k data ($n =  2622$) has three close clusters in $9\\text{-}D$. Two \\gD{} projection from a tour on $9\\text{-}D$ of the model fit with @fig-pbmc-mse a ($a_1 = 0.06$, $b = 616/86 (22, 28)$) shows three-well separated clusters with big separations. On the other hand, the model fit with @fig-pbmc-mse e ($a_1 = 0.06$, $b = 704/227 (22, 32)$) shows three-well separated clusters with small separations. Therefore, @fig-pbmc-mse e is more reasonable than @fig-pbmc-mse a. Also, the fitted models in $9\\text{-}D$ help to see some unobserved patterns of the data: (i) dense points, and (ii) non-linear clusters. Videos of the langevitour animations are available at <https://youtu.be/3VfK3M2gnZM> and <https://youtu.be/Es84bwQcndU> respectively."
#| label: fig-model-pbmc-author-proj
#| fig-pos: H
#| fig-width: 10
#| fig-height: 15
#| out-width: 90%

@@

@fig-mnist-mse compares the fit of six layouts computed using UMAP (b), PHATE (c), TriMAP (d), PaCMAP (e) with default (hyper-)parameter setting and two tSNE runs, one with default (hyper-)parameter setting (a) and the other changing perplexity to $89$ (f). The layouts are reasonably similar in that they all have the observations in a single blob. Some (b, c) have a more curved shape than others. Layout e is the most different having a linear shape, and a single very large outlier. Both a and f have a small clump of points perhaps slightly disconnected from the other points, in the lower to middle right.

@@

We have developed an approach to help assess and compare NLDR layouts, generated by different methods and (hyper-)parameter choice(s). It depends on conceptualising the \gD{} layout as a model, creating a wireframe representation of it that can be lifted into \pD{}. The fit is assessed by viewing the model in the data space, computing residuals and RMSE. Different layouts can be compared using the RMSE, and provides quantitative and objective methods for deciding on the most suitable NLDR layout to represent the \pD{} data. It also provides a way to predict the values of new \pD{} observations in the \gD{}, which could be useful for implementing uncertainty checks such as using training and testing samples.

>>>>

---
title: "Looking at Non-Linear Dimension Reductions as Models in the Data Space"
format: 
    jasa-pdf:
        keep-tex: true
    jasa-html: default
author:
  - name: Jayani P. Gamage
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-6265-6481
    email: jayani.piyadigamage@monash.edu
    url: https://jayanilakshika.netlify.app/
  - name: Dianne Cook
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-3813-7155
    email: dicook@monash.edu 
    url: http://www.dicook.org/
  - name: Paul Harrison
    affiliations:
      - name: Monash University
        department: MGBP, BDInstitute
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-3980-268X
    email: 	paul.harrison@monash.edu
    url: 
  - name: Michael Lydeamore
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0001-6515-827X
    email: michael.lydeamore@monash.edu
    url: https://www.michaellydeamore.com/
  - name: Thiyanga S. Talagala
    affiliations:
      - name: University of Sri Jayewardenepura
        department: Statistics
        address: Gangodawila
        city: Nugegoda 
        country: Sri Lanka
        postal-code: 10100
    orcid: 0000-0002-0656-9789
    email: ttalagala@sjp.ac.lk 
    url: https://thiyanga.netlify.app/
tbl-cap-location: bottom
abstract: |
  Non-linear dimension reduction (NLDR) techniques such as tSNE, and UMAP provide a low-dimensional representation of high-dimensional data (\pD{}) by applying a non-linear transformation. NLDR often exaggerates random patterns, sometimes due to the samples observed. But NLDR views have an important role in data analysis because, if done well, they provide a concise visual (and conceptual) summary of \pD{} distributions. The NLDR methods and (hyper-)~parameter choices can create wildly different representations, making it difficult to decide which is best, or whether any or all are accurate or misleading. To help assess the NLDR and decide on which, if any, is the most reasonable representation of the structure(s) present in the \pD{} data, we have developed an algorithm to show the \gD{} NLDR model in the \pD{} space, viewed with a tour, a movie of linear projections. From this, one can see if the model fits everywhere, or better in some subspaces, or completely mismatches the data. Also, we can see how different methods may have similar summaries or quirks.

@@

We define the NLDR as a function $g\text{:}~ \mathbb{R}^{n\times p} \rightarrow \mathbb{R}^{n\times k}$, with (hyper-)~parameters $\mathbfit{\theta}$. The parameters, $\mathbfit{\theta}$, depend on the choice of $g$, and can be considered part of model fitting in the traditional sense. Common choices for $g$ include functions used in tSNE, UMAP, PHATE, TriMAP, PaCMAP, or MDS, although in theory any function that does this mapping is suitable.

@@

```{r}
#| label: best-fit-tsne
#| fig-cap: "Lifting the fitted model into \\pD{}. Starting from a default tSNE layout (a) of non-linear clusters data ($n =  2000$ and $p = 4$), from which the \\gD{} wireframe is constructed (a) and then lifted to $4\\text{-}D$ (b, c). The fit is reasonbly tight with the data, although it does not fully spread the full width. This highlights a key characteristic of tSNE, which compresses the data when applying tSNE. Furthermore, the sparse space observed in the middle of the tSNE layout is a result of (hyper-)~parameter choice. Video of the langevitour animation is available at <>."
#| fig-width: 8
#| fig-height: 2

@@

@fig-five-gau-projs a1, b1, c1 show the \gD{} layouts for (a) tSNE, (b) UMAP, and (c) PaCMAP, respectively. The default (hyper-)~parameters for each method are used. In each layout we can see an accurate representation where all five clusters are visible, although with varying degrees of separation.

@@

@fig-one-dens_clust-error (a2, b2) shows \gD{} layout for tSNE created using the default (hyper-)~parameters. One would expect to see a rectangular shape if the curved sheet is flattened, but the layout is triangular. The other two displays show the residuals as a dot density plot (a1, b1), and a \gD{} projection of the data and the model from \fD{} (a3, b3). Using linked brushing between the plots, we can highlight points in the tSNE layout, and examine where they fall in the original \fD{}. The darker (maroon) points indicate points that have been highlighted by linking. In row a, the points at the top of the triangle are highlighted, and we can see these correspond to higher residuals, and also to all points at the low density end of the curved sheet. In row b, points at the lower left side of the triangle are highlighted which corresponds to smaller residuals and one corner of the sheet at the high density end of the curved sheet.

@@

```{r}
#| echo: false
#| fig-cap: "Compare the published \\gD{} layout (@fig-pbmc-mse a) and the \\gD{} layout selected (@fig-pbmc-mse e) by MSE plot (@fig-pbmc-mse) from the UMAP and tSNE with default (hyper-)~parameters. The PBMC3k data ($n =  2622$) has three close clusters in $9\\text{-}D$. Two \\gD{} projection from a tour on $9\\text{-}D$ of the model fit with @fig-pbmc-mse a ($a_1 = 0.06$, $b = 616/86 (22, 28)$) shows three-well separated clusters with big separations. On the other hand, the model fit with @fig-pbmc-mse e ($a_1 = 0.06$, $b = 704/227 (22, 32)$) shows three-well separated clusters with small separations. Therefore, @fig-pbmc-mse e is more reasonable than @fig-pbmc-mse a. Also, the fitted models in $9\\text{-}D$ help to see some unobserved patterns of the data: (i) dense points, and (ii) non-linear clusters. Videos of the langevitour animations are available at <https://youtu.be/3VfK3M2gnZM> and <https://youtu.be/Es84bwQcndU> respectively."
#| label: fig-model-pbmc-author-proj
#| fig-pos: H
#| fig-width: 10
#| fig-height: 15
#| out-width: 90%

@@

@fig-mnist-mse compares the fit of six layouts computed using UMAP (b), PHATE (c), TriMAP (d), PaCMAP (e) with default (hyper-)~parameter setting and two tSNE runs, one with default (hyper-)~parameter setting (a) and the other changing perplexity to $89$ (f). The layouts are reasonably similar in that they all have the observations in a single blob. Some (b, c) have a more curved shape than others. Layout e is the most different having a linear shape, and a single very large outlier. Both a and f have a small clump of points perhaps slightly disconnected from the other points, in the lower to middle right.

@@

We have developed an approach to help assess and compare NLDR layouts, generated by different methods and (hyper-)~parameter choice(s). It depends on conceptualising the \gD{} layout as a model, creating a wireframe representation of it that can be lifted into \pD{}. The fit is assessed by viewing the model in the data space, computing residuals and RMSE. Different layouts can be compared using the RMSE, and provides quantitative and objective methods for deciding on the most suitable NLDR layout to represent the \pD{} data. It also provides a way to predict the values of new \pD{} observations in the \gD{}, which could be useful for implementing uncertainty checks such as using training and testing samples.
